---
title: "HaVu_hw4"
author: "Vu Thi-Hong-Ha"
date: "March 13, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse, suppressMessages())
library(ggplot2, suppressMessages())
library(car, suppressMessages())
```

## Question 1:
We have:
$$
X_{n \times p} = A_{n \times (g+1)}\beta_{(g+1) \times p} + \epsilon_{n \times p}
$$
and
$$
\hat{\beta}_{g \times p} = (A'_{g \times n}A_{n \times g})^{-1}A'_{g \times n}X_{n \times g}
$$
Then,
$$
X_k = A\beta_k + \epsilon_k
$$

$$
\hat{\beta}_{k} = (A'A)^{-1}A'X_{k}
$$
Therefore,
$$
\mathbf{E}[ \hat{\beta}_k ] = \mathbf{E}[(A'A)^{-1}A'X_{k}] = (A'A)^{-1}A'\mathbf{E}[X_k]
$$

$$
\mathbf{E}[X_k] = \mathbf{E}[A\beta_k + \epsilon_k] = \mathbf{E}[A\beta_k] + \mathbf{E}[\epsilon_k] = \mathbf{E}[\epsilon_k]
$$
because $A\beta_k$ is a constant matrix so $\mathbf{E}[A\beta_k] = A\beta_k$. <br />
As $\epsilon_k \sim N(0, \Sigma_k)$, $\mathbf{E}[\epsilon_k] = 0$, and $\mathbf{E}[X_k] = A\beta_k$. <br />
Hence,
$$
\mathbf{E}[ \hat{\beta}_k ] = (A'A)^{-1}A'A\beta_k = \beta_k
$$
Next,
$$
Var[\hat{\beta}_k] = Var[(A'A)^{-1}A'X_{k}] = (A'A)^{-1}A'Var(X_{k})[(A'A)^{-1}A']'
$$

$$
Var[X_k] = Var[A\beta_k + \epsilon_k] = Var[\epsilon_k] = \sigma_{kk}
$$
Hence,
$$
Var[\hat{\beta}_k] = (A'A)^{-1}A'\sigma_{kk}[(A'A)^{-1}A']' = (A'A)^{-1}A'[(A'A)^{-1}A']'\sigma_{kk} = (A'A)^{-1}A'A[(A'A)^{-1}]'\sigma_{kk} = \sigma_{kk}(A'A)^{-1}
$$
Next,
$$
cov(\hat{\beta}_k, \hat{\beta}_i) = cov((A'A)^{-1}A'X_{k}, (A'A)^{-1}A'X_{i}) = (A'A)^{-1}A'\mathbf{E}[(X_k - \mathbf{E}(X_k))(X_i - \mathbf{E}(X_i)) ][(A'A)^{-1}A']' = (A'A)^{-1}A'\sigma_{ki}[(A'A)^{-1}A']' = \sigma_{ki}(A'A)^{-1}
$$


## Question 2:
*a)*
```{r}
olive <- read.table("D:/Coursework/Stat 501 Spring 2020/Homework3/olive.dat", header = T)

##a
temp <- olive %>% group_by(group.id) %>% summarise_all(mean)
temp <- temp[,-1]
data <- data.frame(matrix(ncol = 3, nrow = 9*8))
colnames(data) <- c("group.id", "chemical", "sampleMean")
data$group.id <- as.character(c(rep(1, 8), rep(2, 8), rep(3, 8), rep(4, 8), rep(5, 8),
                                rep(6, 8), rep(7, 8), rep(8, 8), rep(9, 8)))
data$chemical <- rep(c("1", "2", "3", "4", "5", "6", "7", "8"), 9)
data$sampleMean <- as.numeric(c(temp[1,], temp[2,], temp[3,], temp[4,], temp[5,], temp[6,],
                                temp[7,], temp[8,], temp[9,]))

ggplot(data = data, aes(x=as.numeric(chemical), y=sampleMean, color=group.id)) + geom_line() +
  xlab("Chemicals") + ylab("Sample Mean")
```
From the plot we can see that some chemicals have similar mean compositions across all regions, but some have very distinctive impacts (chemical X1, X4 and X5, for example). <br />

*b)*
```{r}
olive$region <- NA
for (i in 1:nrow(olive)) {
  if (olive$group.id[i] == "1" | olive$group.id[i] == "2" | olive$group.id[i] == "3" | olive$group.id[i] == "4") {
    olive$region[i] = "R1"
  }
  else if (olive$group.id[i] == "5" | olive$group.id[i] == "6") {
    olive$region[i] = "R2"
  }
  else {
    olive$region[i] = "R3"
  }
}

olive$region <- as.factor(olive$region)

fit <- manova(as.matrix(olive[,2:9]) ~ olive$region)
fit #view model

#two tests for mean differences
summary(fit, test="Wilks")
summary(fit, test="Pillai")
```
Based on the two tests above, we reject the null that the mean components are similar across three regions. <br />


*c)*
```{r}
olive$group.id <- as.factor(olive$group.id)

fit.lm <- lm(cbind(X1, X2, X3, X4, X5, X6, X7, X8)~group.id , data = olive, contrasts = list(group.id = contr.SAS))

C <- matrix(c(0, 1, 0, 0, 0, 0, 0, 0, 0,
              0, 0, 1, 0, 0, 0, 0, 0, 0,
              0, 0, 0, 1, 0, 0, 0, 0, 0,
              0, 0, 0, 0, 1, 0, 0, 0, 0,
              0, 0, 0, 0, 0, 1, 0, 0, 0,
              0, 0, 0, 0, 0, 0, 1, 0, 0,
              0, 0, 0, 0, 0, 0, 0, 1, 0,
              0, 0, 0, 0, 0, 0, 0, 0, 1), ncol = 9,  by = T)
M <- matrix(c(1, 0, 0, 0, 0, 0, 0,
             -1, 1, 0, 0, 0, 0, 0,
             0, -1, 1, 0, 0, 0, 0,
             0, 0, -1, 1, 0, 0, 0,
             0, 0, 0, -1, 1, 0, 0,
             0, 0, 0, 0, -1, 1, 0,
             0, 0, 0, 0, 0, -1, 1,
             0, 0, 0, 0, 0, 0, -1), nrow = 8, by = T)

newfit <- linearHypothesis(model = fit.lm, hypothesis.matrix = C, P = M)
print(newfit)
```
Based on this result, we reject the null that the regions and chemical components have no interaction. Matrix $C$ and $M$ are:
```{r}
C
M
```


*d)*
```{r}
##d
region1 <- subset(olive, olive$region == "R1")

fit.lm.r1 <- lm(cbind(X1, X2, X3, X4, X5, X6, X7, X8)~group.id , data = region1, contrasts = list(group.id = contr.SAS))

C.r1 <- matrix(c(0, 1, 0, 0,
                 0, 0, 1, 0,
                 0, 0, 0, 1), ncol = 4,  by = T)

fit.r1 <- linearHypothesis(model = fit.lm.r1, hypothesis.matrix = C.r1)

print(fit.r1)
```

We reject the null that means of chemical components among the 4 areas within region R1 are equivalent. The matrix $C$ is:
```{r}
C
```
Matrix M is the I matrix.


## Question 3:
*a)* <br />
Looking at the data, we can expect the rate of change is much smaller than 1. Say, in a year, the data is $X_t$, and by the next year, the data $X_{t+1}$changes by $100r\%$. Then $X_{t+1} = X_t(1+r)$. <br />
If $r << 1$, then $ln(1+r) \approx r$. <br />
Then $ln(X_{t+1}) = ln(X_t) + ln(1+r) = ln(X_t) + r$. <br />
Hence, $ln(X_{t+1}) - ln(X_t)$ is approximately the rate of change in data. <br />

*b)*
```{r}
fbiwide <- read.table("D:/Coursework/Stat 501 Spring 2020/Homework4/fbiwide.txt", sep = " ", header = T, row.names = 1)
fbiwide <- subset(fbiwide, fbiwide$State %in% c("California", "Iowa", "Illinois", "District of Columbia", "New York"))
fbiwide <- fbiwide[, !(names(fbiwide) %in% c("Legacy.rape", "Rape")),]
fbiwide_trans <- log(fbiwide[,4:10])
fbiwide[,4:10] <- fbiwide_trans
fbiwide$decade <- NA
d1 <- seq(1961, 1980, by = 1)
d2 <- seq(1981, 2000, by = 1)
d3 <- seq(2001, 2017, by = 1)
for (i in 1:nrow(fbiwide)) {
  if (fbiwide$Year[i] %in% d1) {
    fbiwide$decade[i] <- "1"
  }
  else if (fbiwide$Year[i] %in% d2) {
    fbiwide$decade[i] <- "2"
  }
  else {
    fbiwide$decade[i] <- "3"
  }
}

fbiwide_sub <- fbiwide[, which(names(fbiwide) %in% c("State", "Aggravated.assault", "Burglary",
                                                     "Larceny.theft", "Motor.vehicle.theft",
                                                     "Murder", "Robbery", "decade"))]
fbiwide_sub$State <- as.factor(fbiwide_sub$State)
fbiwide_sub$decade <- as.factor(fbiwide_sub$decade)  

fit.lm <- lm(cbind(Aggravated.assault, Burglary, Larceny.theft, 
                   Motor.vehicle.theft, Murder, Robbery) ~ State * decade, data = fbiwide_sub)

fit.manova <- Manova(fit.lm)

```

The results of two-way MANOVA to test for interaction between states and decades are
```{r, eval=F}
Term: State:decade 

Sum of squares and products for the hypothesis:
                    Aggravated.assault  Burglary Larceny.theft Motor.vehicle.theft    Murder   Robbery
Aggravated.assault           18.121533 14.210256      3.406088            9.666346  5.006994 12.834579
Burglary                     14.210256 16.052879      3.971727           10.033611  6.834359 11.965163
Larceny.theft                 3.406088  3.971727      1.409322            3.652452  1.738742  3.202062
Motor.vehicle.theft           9.666346 10.033611      3.652452           25.284842 11.631135 15.534786
Murder                        5.006994  6.834359      1.738742           11.631135  7.967510  8.379406
Robbery                      12.834579 11.965163      3.202062           15.534786  8.379406 13.905902

Multivariate Tests: State:decade
                 Df test stat  approx F num Df   den Df     Pr(>F)    
Pillai            8  1.938368  15.80853     48 1590.000 < 2.22e-16 ***
Wilks             8  0.051007  22.21457     48 1283.371 < 2.22e-16 ***
Hotelling-Lawley  8  5.315965  28.61023     48 1550.000 < 2.22e-16 ***
Roy               8  3.156224 104.54991      8  265.000 < 2.22e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```


We reject the null that there is no interaction between states and decades. <br />



*c)* <br />
Sum of squares and products for the hypothesis about `States`, `decade` and interactions betwene them is the following:
```{r}
fit.manova$SSP
```

Sum of squares and products for the residuals are:
```{r}
fit.manova$SSPE
```



*d)* <br />
```{r}
igiveup <- data.frame(matrix(ncol = 4, nrow = nrow(fbiwide_sub)))
names(igiveup) <- c("State", "decade", "Aggravated.assault - Burglary", "Burglary - Larceny.theft")
igiveup$State <- fbiwide_sub$State
igiveup$decade <- fbiwide_sub$decade
igiveup$`Aggravated.assault - Burglary` <- fbiwide_sub$Aggravated.assault - fbiwide_sub$Burglary
igiveup$`Burglary - Larceny.theft` <- fbiwide_sub$Burglary - fbiwide_sub$Larceny.theft

fit.lm.lastquestion <- lm(cbind(`Aggravated.assault - Burglary`, `Burglary - Larceny.theft`) ~ State * decade, data = igiveup)

fit.manova.lastquestion <- Manova(fit.lm.lastquestion)

summary(fit.manova.lastquestion)
```

